Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-03 tex/draft.tex
First edited draft

\newpage
\tableofcontents
\newpage
\doublespacing
%\linenumbers
\pagenumbering{arabic}

@@

    All SM particles transform according to the gauged symmetry group\fxnote{be good at describing the meaning of `gauged symmetry group'} $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are eight gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into\fxnote{make sure I'm good at explaining this}. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin~\cite{sm_intro}.

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV)\fxnote{maybe just use $v$} of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $y$~\cite{sm_intro}.

@@

\section{Naturalness and the Higgs mass}
    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry~\cite{giudice_naturally, thooft_naturalness}. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory. \fxnote{wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that also explains cases where dimensional analysis fails.

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$~\cite{dine_naturalness}\fxnote{show diagrams and terms?}. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}~\cite{cms_higgs, atlas_higgs}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important~\cite{giudice_naturally}. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work.

    As a fundamental scalar, the Higgs is doubly cursed. As explained in section \ref{SM}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms~\cite{giudice_naturally}.\fxnote{wordy as hell}

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass~\cite{craig}. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass itself. Finally, the dominant\fxnote{what do I mean by dominant?} BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below~\cite{dine_naturalness}.

@@

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the lack of sparticles in collider searches implies that SUSY must be broken. With broken SUSY, the same diagram as above produces a correction of the form~\cite{feng}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

@@

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional SUSY searches~\cite{primer}.

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs~\cite{fate}. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present\fxnote{say better}. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below~\cite{cms_experiment}.

@@

    \subsection{Tracker}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupacy numbers}

@@

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the $\num{20} < r < \SI{116}{cm}$ and $\lvert z \rvert < \SI{282}{cm}$ region of CMS. Composed of \num{9.3} million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly \SI{196}{m^2}. At each trigger, the strip output voltage is read out via optical link~\cite{cms_pixel_upgrade}.

    \subsection{Electromagnetic Calorimeter}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}.

@@

        The barrel section (HB) spans a pseudorapidty range of \num{-1.4} to \num{1.4} and has \num{2304} towers \fxnote{check}, each of which contains \num{14} \fxnote{check} brass plates and covers an area of \num{0.087} x \SI{0.087}{in} $\eta-\phi$ \fxnote{get a feel for this area}. In addition, an extra layer (or two at $\eta = \num{0}$)of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidty range of \num{-1.26} to \num{1.26} and increases the minimum effective HCAL interaction length to greater than \num{11.8}.

        Each endcap spans a pseudorapidity range of \num{1.3} to \num{3.0} with \num{14} towers in eta and \num{5} to \SI{10}{\degree} $\phi$ segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits \SI{11.2}{m} from the interaction pipe in a pseudorapidty range of \num{3.0} to \num{5.0}. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. \fxnote{say why HF is useful or mention crazy flux?}

@@

    \subsection{Muon System}
        \fxnote{Specifically mention CMS design specs to meet physics goals?}
        The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon pT\fxnote{define} resolution at high momenta where the tracker $\pT$ resolution declines \fxnote{quantify or show std plot?}\cite{cms_experiment, cms_tdr}.

        In the barrel detector (BD), four layers of DTs and and RPCs are embedded in the solenoid return yoke at \num{4.0}, \num{4.9}, \num{5.9}, and \SI{7.9}{m}. The maximum drift length is \SI{2}{\cm} and the single point resolution is \SI{200}{\cm}. When reconstructing a vector describing the muon momentum, the $\phi$ precision is better than \SI{100}{\micro\m} in position and \SI{1}{\milli\radian} in direction.

        Each muon endcap detector (ME) is composed of \num{234} CSCs and four RPC stations \fxnote{figure out exactly what RPC station means and say that instead}. Each CSC has six gas gaps with radial cathode strips and nearly perpendicualar planes of anode wires. The wire signal from the ionization-induced electron avalanch is fast enough to use in the L1 trigger, while the slower \fxnote{quantify} cathode strip signal provides precise \fxnote{quantify} position measurements. All in all, the CSC spacial resolution is approximately \SI{200}{\micro\m} and the angular resolution is approximately \SI{10}{\milli\radian}.

    \subsection{Trigger}
        \fxnote{Is entire detector read out at L1?}
        The trigger reduces the data writing rate from the \SI{40}{\mega\hertz} collision rate to less than \SI{1}{\kilo\hertz} so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects ($\mathrm{e}$/$\gamma$, $\tau$, jets, $\mu$, and $\MET$) and uses custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz\cite{cms_experiment, cms_trigger_upgrade}.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/jamie_cartoon}
        \caption{Diagram of several LLP topologies. Image curtesy of Jamie Antonelli}
        \label{jamie_cartoon}
        \end{center} \end{figure}

@@

        Slow and stopped particle searches, on the other hand, are sensitive to detector-stable\fxnote{define?} LLPs that are far more massive\fxnote{be able to quantify} than any stable SM particle. The lower velocity of these high-mass particles shows up as increased ionization energy loss per unit distance ($\langle\frac{dE}{dx}\rangle$) or longer time-of-flight (TOF) measurements than SM particles, which generally have $\beta > \num{0.9}$\fxnote{check this number}~\cite{cms_hscp}. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy\fxnote{how long and how small?}. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings~\cite{cms_stopped}.

        These\fxnote{I've been saying `these' a lot recently} four approaches cover a wide\fxnote{how wide} range of LLP lifetimes that would be missed in traditional searches. The next section will look  at a recent example of each analysis type from either ATLAS or CMS.

@@

        This search\fxnote{repetative wording} uses a simplified benchmark model in the Split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in Fig.~\ref{displaced_process}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measureable distance before decaying.

@@

        Jets and $\MET$ are reconstructed as in conventional analyses, but dispaced vertices demand special care when identifying tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters ($\mathrm{d}_0$ and $\mathrm{z}_0$, respectively) of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing requirements on $\mathrm{d}_0$, $\mathrm{z}_0$, and number of hits, LRT is able to reconstruct tracks that point back to DVs.

@@

    \noindent \begin{equation}
        \Delta_i = \frac{\partial \ln{\mHiggs^2}}{\partial \ln{\alpha_i}} \quad ,
        \label{fine_tuning}
    \end{equation}

>>>>

\newpage
\tableofcontents
\newpage
\doublespacing
\linenumbers
\pagenumbering{arabic}

@@

    All SM particles transform according to the gauged symmetry group\fxnote{be good at describing the meaning of `gauged symmetry group'} $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are eight gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are categorized by which representation of each group they fit into\fxnote{make sure I'm good at explaining this}. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distinguishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where $Q$ is electric charge and $I_3$ is the third component of weak isospin~\cite{sm_intro}.

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV)\fxnote{maybe just use $v$} of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $y$~\cite{sm_intro}.

@@

\section{Naturalness and the Higgs mass}
    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry~\cite{giudice_naturally, thooft_naturalness}. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory\fxnote{wordy and maybe repetitive}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that also explains cases where dimensional analysis fails.

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$~\cite{dine_naturalness}\fxnote{show diagrams and terms?}. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}~\cite{cms_higgs, atlas_higgs}, a value that must be equal to the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important~\cite{giudice_naturally}. While it's possible that the bare mass and all corrections just so happen to cancel so perfectly, such miraculous fine tuning could also be a hint that a deeper physical mechanism is at work.

    As a fundamental scalar, the Higgs is doubly cursed. As explained in section \ref{SM}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms~\cite{giudice_naturally}.\fxnote{confusing paragraph. Rewrite and add math}

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass~\cite{craig}. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass. Finally, the dominant\fxnote{what do I mean by dominant?} BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below~\cite{dine_naturalness}.

@@

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the uniformly null results in collider searches imply SUSY must be broken. With broken SUSY, the sum of diagrams in Fig.~\ref{exact_susy_correction} produces a correction of the form~\cite{feng}

@@

    The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenomena. A handful of models with particular approaches to SUSY breaking that are particularly prominent or particularly relevant to naturalness or long-lived particles are gauge-mediated SUSY breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating (RPV) SUSY. The details of these models will be explored as needed throughout this paper.

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that would be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay before reaching the inner tracker to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

@@

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-parity, a proposed symmetry that prevents proton decay~\cite{primer}\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenological consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional SUSY searches~\cite{primer}.

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs~\cite{fate}. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which suppresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 through to the present\fxnote{cite something}. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}~\cite{cms_experiment}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below.

@@

    \subsection{Tracker}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupancy numbers}

@@

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the $\num{20} < r < \SI{116}{cm}$ and $\lvert z \rvert < \SI{282}{cm}$ region of CMS. Composed of \num{9.3} million silicon strips that vary in size \fxnote{look up and specify} depending on particle flux in the region, the SST covers an active area of roughly \SI{196}{m^2}. At each trigger, the strip output voltage is read out via optical link~\cite{cms_pixel_upgrade}.

    \subsection{Electromagnetic Calorimeter}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tungstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{\nano\s}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necessitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}.

@@

        The barrel section (HB) spans a pseudorapidity range of \num{-1.4} to \num{1.4} and has \num{2304} towers \fxnote{check}, each of which contains \num{14} \fxnote{check} brass plates and covers an area of \num{0.087} x \SI{0.087}{in} $\eta-\phi$ \fxnote{get a feel for this area}. In addition, an extra layer (or two at $\eta = \num{0}$) of scintillator tiles sits just outside the solenoid. This extra layer, known as hadron outer (HO), spans a pseudorapidity range of \num{-1.26} to \num{1.26} and increases the minimum effective HCAL interaction length to greater than \num{11.8}.

        Each endcap spans a pseudorapidity range of \num{1.3} to \num{3.0} with \num{14} towers in eta and \num{5} to \SI{10}{\degree} $\phi$ segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits \SI{11.2}{m} from the interaction pipe in a pseudorapidity range of \num{3.0} to \num{5.0}. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. \fxnote{say why HF is useful or mention crazy flux?}

@@

    \subsection{Muon System}
        \fxnote{Specifically mention CMS design specs to meet physics goals?}
        The muon system (MS), which is the outermost subdetector, detects muons with \num{3} varieties of gaseous ionization detectors: drift tubes (DTs) in the barrel, cathode strip chambers (CSCs) in the endcaps, and resistive plate chambers (RPCs) in both the barrel and endcaps. The choice of detector varies between barrel and endcap due to the difference in neutron-induced background\fxnote{what?} , muon rates, and magnetic field strength\fxnote{why does B matter?}. The RPCs are useful in the endcaps and barrel because their fast response and good time resolution allow unambiguous bunch crossing identification. Combining MS and tracker measurements improves the muon momentum resolution at high momenta where the tracker resolution declines \fxnote{quantify or show std plot?}\cite{cms_experiment, cms_tdr}.

        In the barrel detector (BD), four layers of DTs and RPCs are embedded in the solenoid return yoke at \num{4.0}, \num{4.9}, \num{5.9}, and \SI{7.9}{m}. The maximum drift length is \SI{2}{\cm} and the single point resolution is \SI{200}{\cm}. When reconstructing a vector describing the muon momentum, the $\phi$ precision is better than \SI{100}{\micro\m} in position and \SI{1}{\milli\radian} in direction.

        Each muon endcap detector (ME) is composed of four layers of CSCs and three layers of RPCs. Each CSC has six gas gaps with radial cathode strips and nearly perpendicular planes of anode wires. The wire signal from the ionization-induced electron avalanche is fast enough to use in the L1 trigger, while the slower \fxnote{quantify} cathode strip signal provides precise \fxnote{quantify} position measurements. All in all, the CSC spacial resolution is approximately \SI{200}{\micro\m} and the angular resolution is approximately \SI{10}{\milli\radian}.

    \subsection{Trigger}
        \fxnote{Is entire detector read out at L1?}
        The trigger reduces the data writing rate from the \SI{40}{\mega\hertz} collision rate to less than \SI{1}{\kilo\hertz} so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 analyzes input from ECAL, HCAL, and MS with custom electronics to reduce the rate to $\approx$100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired $<$ 1 kHz\cite{cms_experiment, cms_trigger_upgrade}.

@@

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/jamie_cartoon}
        \caption{Diagram of several LLP topologies. Image courtesy of Jamie Antonelli}
        \label{jamie_cartoon}
        \end{center} \end{figure}

@@

        Slow and stopped particle searches, on the other hand, are sensitive to detector-stable\fxnote{define?} LLPs that are far more massive\fxnote{be able to quantify} than any stable SM particle. The lower velocity of these high-mass particles can show up as increased ionization energy loss per unit distance ($\langle\frac{dE}{dx}\rangle$) or longer time-of-flight (TOF) measurements than SM particles, which generally have $\beta > \num{0.9}$\fxnote{check this number}~\cite{cms_hscp}. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy\fxnote{how long and how small?}. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings~\cite{cms_stopped}.

        These\fxnote{I've been saying `these' a lot recently} four approaches cover a wide\fxnote{how wide} range of LLP lifetimes that would be missed in traditional searches. The next sections will examine a recent example of each analysis type from either ATLAS or CMS.

@@

        This search\fxnote{repetitive wording} uses a simplified benchmark model in the Split SUSY paradigm where gluinos are kinematically accessible but squarks are not. The relevant process, shown in Fig.~\ref{displaced_process}, is pair-produced gluinos that decay to a quark-\fxnote{how to punctuate?}virtual squark pair. The virtual squark then decays to a quark and the LSP, presumed to be a neutralino\fxnote{have I defined neutralino?}\fxnote{why is this a reasonable assumption?}. The large mass of the intermediate squark suppresses the gluino decay, giving the gluino enough time to form an R-hadron\fxnote{first use. Define here if not in LLP section} that propagates some measurable distance before decaying.

@@

        Jets and $\MET$ are reconstructed as in conventional analyses, but displaced vertices demand special care when identifying tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters ($\mathrm{d}_0$ and $\mathrm{z}_0$, respectively) of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing requirements on $\mathrm{d}_0$, $\mathrm{z}_0$, and number of hits, LRT is able to reconstruct tracks that point back to DVs.

@@

    \noindent \begin{equation}
        \Delta_i = \frac{\partial \ln{\mHiggs^2}}{\partial \ln{\alpha_i}} ,
        \label{fine_tuning}
    \end{equation}

    \noindent where $\alpha_i$ is a particular model parameter and $\Delta_i$ is the degree of fine tuning associated with that parameter. The overall degree of fine tuning for the model can then be assigned as $\max{\left(\Delta_i\right)}$~\cite{cornering}.
